Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model

### 2022-07-25 Rmarkdown/paper.Rmd
edits to conclusion

We demonstrate our approach using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols. Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary.

@@

Time series of particulate and dissolved lead featured prominent negative trends after the introduction of orthophosphate, followed by an extended period of relative stability. The dissolved lead time series were seasonal, with maximum levels in late summer and minimum levels in early spring (Figure 4a). To estimate the time-varying mean representing each series, we modeled lead as the sum of a global multi-year smoother, a seasonal smoother (dissolved lead only), a group-level multi-year smoother, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).

@@

Accounting for autocorrelation also improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation. Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016]. We applied it to the particulate and dissolved lead models with continuous-time autoregressive errors and also to identically-specified models with independent errors. The autoregressive error models yielded higher expected log pointwise predictive densities (i.e., greater accuracy), exceeding the independent error models in this metric by 206 and 47 (dissolved and particulate lead models, respectively).

Posterior predictions from the models---$\mu_t$ in Equation $(5)$---can be compared to contrast the orthophosphate dosing protocols. Here we compared the 0--0.5 mg P L^-1^ with the 1 mg P L^-1^ and the 1 mg P L^-1^ with the 2--0.75 mg P L^-1^ dosing protocols over time by considering their ratios.

The three protocols yielded distinct lead release distributions over most of the study, and the difference between the lowest orthophosphate dose (0.5 mg P L^-1^) and the other two is particularly apparent (Figure 7). The ratio of dissolved lead concentrations in the absence of orthophosphate to those representing the 1 mg P L^-1^ dose peaked at `r signif(fig7_summ$.epred, 3)` with a 95% credible interval of `r signif(fig7_summ$.lower, 3)`--`r signif(fig7_summ$.upper, 3)` (Figure 7a); the ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^, but its 2.5th percentile---the lower bound of the 95% credible interval---remained above 1 for the majority of the study. This pattern is consistent with expectations based on geochemical solubility models [@schock1996corr].

The ratio comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b). As a comparison of either particulate or dissolved lead concentrations, it was less than one over much of the first part of the series and greater than one over much of the latter part. The transition occurred a short time after orthophosphate was decreased from 2--0.75 mg P L^-1^, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose, likely as particles and small colloids [@trueman_characterizing_2019; @li_impact_2020]. Meanwhile, this dose presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^. This pattern implies a potential advantage and drawback of applying a high passivation dose, followed by a lower maintenance dose, of orthophosphate. That is, a passivation dose may cause excess lead release in the short term---likely due to particle-generating mechanisms discussed elsewhere [@lytle_lead_2020; @zhao_formation_2018]---while ensuring that a lower maintenance dose is effective.

@@

Pipe loop study data can be difficult to analyze, owing to non-linear time trends, seasonality, autocorrelation due to repeated measurements of pipe effluent, irregular sampling frequency, extreme particulate metals concentrations, and left-censoring, or values below the detection limit. To address these difficulties, we propose a robust Bayesian generalized additive model with continuous-time first-order autoregressive errors. With this approach, there is no need to impute non-detects, assume linearity, or treat outliers specially. Moreover, the model accounts for the reduced information content in autocorrelated observations. We demonstrate it using data from a four-year pipe loop study and we use the model to compare orthophosphate dosing protocols and analyze time trends. The additive smooth terms comprising the model yield an informative decomposition of pipe loop time series and a more nuanced comparison of corrosion control treatments.

The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the ratio difference between the 0 and 1 mg P L^-1^ dose. The effect of orthophosphate dose was complex: while the 1 mg P L^-1^ dose resulted in significantly lower lead release than the 0.5 mg P L^-1^ dose, the intermediate dose (0.75 mg P L^-1^) often yielded lower predicted lead levels in the latter half of the study, which we attribute to an initial high dose of 2 mg P L^-1^. Consistent with previous work, orthophosphate tended to increase the particulate fraction of total lead. The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring. This seasonal pattern may have multiple drivers: iron and water temperature, for instance, were also seasonal.

>>>>

We demonstrate the model using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols. Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary.

@@

Time series of particulate and dissolved lead featured prominent negative trends after the introduction of orthophosphate, followed by an extended period of relative stability (Figure 6). The dissolved lead time series were seasonal, with maximum levels in late summer and minimum levels in early spring (Figure 4a).

Posterior predictions from the models---$\mu_t$ in Equation $(5)$---can be compared to evaluate the orthophosphate dosing protocols. Here we compared the 0--0.5 mg P L^-1^ with the 1 mg P L^-1^ and the 1 mg P L^-1^ with the 2--0.75 mg P L^-1^ dosing protocols over time by considering their ratios.

The three dosing protocols yielded distinct lead release distributions over most of the study, and the difference between the lowest orthophosphate dose and the other two is particularly apparent (Figure 7). The ratio of dissolved Pb concentrations at a P ratio of 0:1---comparing the 0 mg P L^-1^ control with the 1 mg P L^-1^ dose---peaked at `r signif(fig7_summ$.epred, 3)` with a 95% credible interval of `r signif(fig7_summ$.lower, 3)`--`r signif(fig7_summ$.upper, 3)` (Figure 7a). The dissolved Pb ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^ (i.e., at a 0.5:1 P ratio), but its 2.5th percentile---the lower bound of the 95% credible interval---remained above 1 with a single brief exception. This highlights the superior corrosion control performance of 1 mg P L^-1^ over both 0 and 0.5 mg P L^-1^ and is consistent with expectations based on geochemical solubility models [@schock1996corr].

Pb ratios comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b). As a comparison of either particulate or dissolved lead concentrations, the Pb ratio at a 1:2 P ratio was less than one, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose. This likely occurred in the dissolved fraction as small colloids [@trueman_characterizing_2019; @li_impact_2020; @lytle_lead_2020].

The transition from 2--0.75 mg P L^-1^ reversed this pattern: the dissolved and particulate Pb ratios at the 1:0.75 P ratio were largely greater than one. A high dose of 2 mg P L^-1^, then, presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^. This implies a potential advantage and drawback of a high passivation dose, followed by a lower maintenance dose, of orthophosphate. That is, a passivation dose may cause excess lead release in the short term---likely due to particle-generating mechanisms [@lytle_lead_2020; @zhao_formation_2018]---while ensuring that a lower maintenance dose is either more effective or effective sooner than it otherwise might be.

## Modeling autocorrelation and assessing performance

@@

Accounting for autocorrelation also improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation on the complete dataset. Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016]. We applied it to the particulate and dissolved lead models with continuous-time autoregressive errors and also to identically-specified models with independent errors. The autoregressive error models yielded higher expected log pointwise predictive densities (i.e., greater accuracy), exceeding the independent error models in this metric by 206 and 47 (dissolved and particulate lead models, respectively).

@@

Pipe loop study data can be difficult to analyze, owing to non-linear time trends, seasonality, autocorrelation due to repeated measurements of pipe effluent, irregular sampling frequency, extreme particulate metals concentrations, and left-censoring, or values below the detection limit. To address these difficulties, we propose a robust Bayesian generalized additive model with continuous-time first-order autoregressive errors. With this approach, there is no need to impute non-detects, assume linearity, or treat outliers specially. Moreover, the model accounts for the reduced information content in autocorrelated observations.

Our approach can be modified to suit the needs of diverse analyses. While time is the sole independent variable here, water quality parameters can easily be added as predictors if inferences concerning their effects are of interest. Moreover, any parameter in the likelihood can be modeled---including the standard deviation if model residuals are heteroskedastic. The model is equally useful for other problems featuring the partially-censored, irregularly-sampled, autocorrelated time series common in water quality analysis [@helsel_statistical]. But like many Bayesian models, this one is computationally-intensive to fit and would be difficult to implement in real-time applications.

We demonstrate it here using data from a four-year pipe loop study and we use the model to compare orthophosphate dosing protocols and analyze time trends. The additive smooth terms comprising the model yield an informative decomposition of pipe loop time series and a more nuanced comparison of corrosion control treatments.

The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the large difference in lead release between the 0 and 1 mg P L^-1^ dose. The effect of orthophosphate dose was complex: lead release at 1 mg P L^-1^ was lower than at 0.5 mg P L^-1^ but higher than at 0.75 mg P L^-1^. We attribute the superior performance of the 0.75 mg P L^-1^ dose to an initial high dose of 2 mg P L^-1^. On the other hand, the initial 2 mg P L^-1^ dose caused excess lead release, highlighting an important trade-off that should be considered in applying a passivation-maintenance (high-low) dosing strategy. And since high orthophosphate doses may generate small lead-orthophosphate colloids/nanoparticles that are difficult to remove from drinking water [@lytle_lead_2020], the 1 mg P L^-1^ dose is the conservative choice. The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring. This seasonal pattern may have multiple drivers: iron and water temperature, for instance, were also seasonal.
